---
title: "BleoTox Radiomics"
author: "C.L.C. Smith"
date: "2023-08-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries needed for this pipeline
Load required packages

```{r , echo=FALSE}
required_packages <- c(
  "tidyverse",
  "caret",
  "readxl",
  "randomForest",
  "pROC",
  "ggplot2"
)

lapply(required_packages, library, character.only = TRUE)
```

# 1) Load data
Load df and create the baseline and interim df

In the df you find the Radiomics values for patients who did and who did not develop toxicity after treatment with a specific type of chemotherapy. 

Radiomics values are for both patient groups extracted for the baseline PET and interim PET

```{r , echo=TRUE}

# ----------------------------------------------------------
# 1. Load data
# ----------------------------------------------------------
df <- read_excel("your_data.xlsx")

# Split the data in the 2 timepoints (Baseline & Interim)
df_baseline <- df %>% filter(Time == "Baseline")
df_interim  <- df %>% filter(Time == "Interim")
```

# 2) Feature selection Volume & SUVmean
All high correlating features (R > 0.7) with either Volume or SUVmean needs to excluded
This is done for both the baseline as interim scans

The feature was removed in all df's if it either correlated with the baseline or interim scans

```{r , echo=TRUE}
# ----------------------------------------------------------
# 2. Feature selection high correlating with Volume & SUVmean
# ----------------------------------------------------------
# Radiomic features with a Spearman correlation coefficient
# R > 0.7 with Volume or SUVmean were excluded.
#
# Spearman correlation was selected because of the
# relatively small sample size and non-parametric nature
# of the data.
# ----------------------------------------------------------

# create new df's
cor_volume_baseline <- vector()
for (i in 1:length(df_baseline)) {
  cor_volume_baseline[i] <- cor(
    df_baseline$Volume, df_baseline[, i], method = "spearman"
  )
}

cor_SUVmean_baseline <- vector()
for (i in 1:length(df_baseline)) {
  cor_SUVmean_baseline[i] <- cor(
    df_baseline$SUVmean, df_baseline[, i], method = "spearman"
  )
}

cor_volume_interim <- vector()
for (i in 1:length(df_interim)) {
  cor_volume_interim[i] <- cor(
    df_interim$Volume, df_interim[, i], method = "spearman"
  )
}

cor_SUVmean_interim <- vector()
for (i in 1:length(df_interim)) {
  cor_SUVmean_interim[i] <- cor(
    df_interim$SUVmean, df_interim[, i], method = "spearman"
  )
}

# Add correlation vectors to dataframe
correlation_df_baseline <- rbind(
  df_baseline, cor_volume_baseline, cor_SUVmean_baseline
)

correlation_df_interim <- rbind(
  df_interim, cor_volume_interim, cor_SUVmean_interim
)

# Extract rows containing correlation coefficients
correlation_rows_baseline <- correlation_df_baseline[
  c(nrow(correlation_df_baseline) - 1, nrow(correlation_df_baseline)),
]

correlation_rows_interim <- correlation_df_interim[
  c(nrow(correlation_df_interim) - 1, nrow(correlation_df_interim)),
]

# Define correlation threshold
cor_threshold <- 0.7

# Identify features with high positive or negative correlations
selected_columns_baseline <- which(
  apply(
    correlation_rows_baseline,
    2,
    function(col) any(abs(col) > cor_threshold)
  )
)

selected_columns_interim <- which(
  apply(
    correlation_rows_interim,
    2,
    function(col) any(abs(col) > cor_threshold)
  )
)

# Remove highly correlated features
filtered_data_baseline <- correlation_df_baseline[, -selected_columns_baseline]
filtered_data_interim <- correlation_df_interim[, -selected_columns_interim]

# Remove appended correlation rows
filtered_data_baseline <- filtered_data_baseline[
  -c(nrow(filtered_data_baseline) - 1, nrow(filtered_data_baseline)),
]

filtered_data_interim <- filtered_data_interim[
  -c(nrow(filtered_data_interim) - 1, nrow(filtered_data_interim)),
]

# Keep only features that remained after filtering
# in both baseline and interim datasets
common_columns <- intersect(names(filtered_data_baseline), names(filtered_data_interim))

# Create new data frames with the common columns
filtered_data_baseline <- filtered_data_baseline[common_columns]
filtered_data_interim <- filtered_data_interim[common_columns]

```
 
# 3) Reduce multicollinearity  
To reduce multicollinearity all high correlating features needs to be excluded
Feature that will be excluded is the feature with the highes overall correlation

In this way we only have features left which do not correlate with each other

```{r , echo= T}

# ----------------------------------------------------------
# 3. Feature selection:
# remove highly intercorrelated radiomic features
# ----------------------------------------------------------
# To reduce multicollinearity, highly correlated
# radiomic features (R > 0.7) were identified.
#
# For correlated feature pairs, the feature with the
# highest overall correlation burden was removed.
# ----------------------------------------------------------

# Create absolute Spearman correlation matrices
cor_matrix_baseline <- abs(cor(filtered_data_baseline, method = "spearman"))
cor_matrix_interim <- abs(cor(filtered_data_interim, method = "spearman"))

# Define correlation threshold
cor_threshold <- 0.7

# ----------------------------------------------------------
# BASELINE
# ----------------------------------------------------------

# Identify highly correlated feature pairs
high_corr_indices_baseline <- which(
  cor_matrix_baseline >= cor_threshold &
    cor_matrix_baseline != 1,
  arr.ind = TRUE
)

# Store features to remove
remove_features_baseline <- c()

for (i in 1:nrow(high_corr_indices_baseline)) {
  
  row <- high_corr_indices_baseline[i, 1]
  col <- high_corr_indices_baseline[i, 2]
  
  # Compare total correlation burden
  row_sum <- sum(cor_matrix_baseline[row, ])
  col_sum <- sum(cor_matrix_baseline[col, ])
  
  # Remove feature with highest correlation burden
  if (row_sum > col_sum) {
    
    remove_features_baseline <- c(
      remove_features_baseline,
      rownames(cor_matrix_baseline)[row]
    )
    
  } else {
    
    remove_features_baseline <- c(
      remove_features_baseline,
      colnames(cor_matrix_baseline)[col]
    )
  }
}

# Remove duplicate feature names
remove_features_baseline <- unique(remove_features_baseline)

# Keep remaining features
selected_data_baseline <- filtered_data_baseline %>%
  select(-all_of(remove_features_baseline))

# ----------------------------------------------------------
# INTERIM
# ----------------------------------------------------------

high_corr_indices_interim <- which(
  cor_matrix_interim >= cor_threshold &
    cor_matrix_interim != 1,
  arr.ind = TRUE
)

remove_features_interim <- c()

for (i in 1:nrow(high_corr_indices_interim)) {
  
  row <- high_corr_indices_interim[i, 1]
  col <- high_corr_indices_interim[i, 2]
  
  row_sum <- sum(cor_matrix_interim[row, ])
  col_sum <- sum(cor_matrix_interim[col, ])
  
  if (row_sum > col_sum) {
    
    remove_features_interim <- c(
      remove_features_interim,
      rownames(cor_matrix_interim)[row]
    )
    
  } else {
    
    remove_features_interim <- c(
      remove_features_interim,
      colnames(cor_matrix_interim)[col]
    )
  }
}

remove_features_interim <- unique(remove_features_interim)

selected_data_interim <- filtered_data_interim %>%
  select(-all_of(remove_features_interim))

# ----------------------------------------------------------
# Keep only overlapping features between datasets
# ----------------------------------------------------------

common_columns <- intersect(
  names(selected_data_baseline),
  names(selected_data_interim)
)

data_baseline_no_high_cor <- selected_data_baseline[
  common_columns
]

data_interim_no_high_cor <- selected_data_interim[
  common_columns
]
```

# 4) Random forest Interim results
We now train a random forest based on the remaining features. We start with a RF for the interim dataset as here the patients already developed toxicity. Now we can see if patients can be classified correctly when they have active toxicity

To do this we again have to add the features Volume and SUVmean to the other remaining features

```{r , echo=FALSE}

# ----------------------------------------------------------
# 4. Random Forest classification interim
# ----------------------------------------------------------
# A random forest classifier was trained to determine
# whether interim radiomic features could distinguish
# patients with acute pulmonary toxicity (DIILD)
# from the same patient group without pulmonary toxicity.
#
# To reduce multicollinearity, only features with low
# correlation to Volume, SUVmean, and other radiomic
# features were retained.
#
# Volume and SUVmean were subsequently added back as
# clinically relevant reference variables.
#
# Due to the limited sample size, leave-one-out
# cross-validation (LOOCV) was applied during training.
# ----------------------------------------------------------

# ----------------------------------------------------------
# Create dataset for patients with toxicity
# ----------------------------------------------------------

forest_toxicity_interim <- data_interim_no_high_cor

# Add clinically relevant variables
forest_toxicity_interim$Volume <- df_interim$Volume
forest_toxicity_interim$SUVmean <- df_interim$SUVmean

# Add outcome label
forest_toxicity_interim$DIILD <- "yes"

# ----------------------------------------------------------
# Create dataset for patients without toxicity
# ----------------------------------------------------------

# Load control dataset
df_control <- read_excel("your_control_data.xlsx")
df_control <- data.frame(df_control)

# Split into baseline and interim scans
Time <- df_control$PatientID

df_control_interim <- df_control %>%
  filter(Time == "Interim")

# Select the same features as toxicity dataset
common_columns <- intersect(names(forest_toxicity_interim),names(df_control_interim))
forest_no_toxicity_interim <- df_control_interim[common_columns]

# Add outcome label
forest_no_toxicity_interim$DIILD <- "no"

# ----------------------------------------------------------
# Combine datasets
# ----------------------------------------------------------

forest_interim <- rbind(forest_toxicity_interim, forest_no_toxicity_interim)
forest_interim$DIILD <- as.factor(forest_interim$DIILD)
str(forest_interim)

# ----------------------------------------------------------
# Random forest model
# ----------------------------------------------------------

# Define tuning grid
tune_grid <- expand.grid(
  .mtry = 1:(ncol(forest_interim) - 1)
)

# Define number of trees to evaluate
ntree_values <- c(
  50, 100, 150, 200,
  250, 300, 500, 1000
)

# Leave-one-out cross-validation
control <- trainControl(method = "LOOCV")

# rf model
rf_results_interim <- list()
for (ntree in ntree_values) {

  set.seed(42)

  rf_model_interim <- train(
    DIILD ~ .,
    data = forest_interim,
    method = "rf",
    trControl = control,
    tuneGrid = tune_grid,
    ntree = ntree
  )

  rf_model_interim$ntree <- ntree
  rf_results_interim[[as.character(ntree)]] <- rf_model_interim
}

# Select best performing model
best_rf_model_interim <- rf_results_interim[[
  which.max(
    sapply(
        rf_results_interim,
        function(x) max(x$results$Accuracy)
      )
    )
  ]]

cat(
  "Best ntree value:",
  best_rf_model_interim$ntree,
  "\n"
)

# Plot model performance
plot(best_rf_model_interim)

# ----------------------------------------------------------
# Feature importance
# ----------------------------------------------------------

final_rf_model_interim <- best_rf_model_interim$finalModel
importance_scores_interim <- importance(final_rf_model_interim)

importance_df_interim <- data.frame(
  Feature = rownames(importance_scores_interim),
  Importance = importance_scores_interim[
    , "MeanDecreaseGini"
  ]
)

ggplot(importance_df_interim, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Feature Importance", x = "Feature", y = "Importance (Mean Decrease in Gini)")


# ----------------------------------------------------------
# Multidimensional scaling (MDS) plot
# ----------------------------------------------------------

oob_predictions_interim <- predict(
  final_rf_model_interim,
  forest_interim,
  type = "prob"
)[, 2]

distance_matrix_interim <- dist(oob_predictions_interim)

mds_result_interim <- cmdscale(distance_matrix_interim, k = 2)

mds_df_interim <- data.frame(
  MDS1 = mds_result_interim[, 1],
  MDS2 = mds_result_interim[, 2],
  Actual = forest_interim$DIILD
)

ggplot(mds_df_interim, aes(x = MDS1, y = MDS2, color = (Actual))) +
  geom_point() +
  labs(title = "MDS Plot of Random Forest Predictions", 
       x = "MDS Dimension 1", 
       y = "MDS Dimension 2", 
       color = "Group"
       )



```


# Random forest Baseline results
We now train a random forest based on the same remaining features but then from the baseline values. This examines if these features can also predict if patients will developed toxicity during chemotherapy.

To do this we again have to add the features Volume and SUVmean to the other remaining features

```{r , echo= True}

# ----------------------------------------------------------
# 4. Random Forest classification
# ----------------------------------------------------------
# A random forest classifier was trained to determine
# whether baseline radiomic features can predict if 
# patients will develop acute pulmonary toxicity 
#
# To reduce multicollinearity, only features with low
# correlation to Volume, SUVmean, and other radiomic
# features were retained.
#
# Volume and SUVmean were subsequently added back as
# clinically relevant reference variables.
#
# Due to the limited sample size, leave-one-out
# cross-validation (LOOCV) was applied during training.
# ----------------------------------------------------------

# ----------------------------------------------------------
# Create dataset for patients with toxicity
# ----------------------------------------------------------

forest_toxicity_baseline <- data_baseline_no_high_cor

# Add clinically relevant variables
forest_toxicity_baseline$Volume <- df_baseline$Volume
forest_toxicity_baseline$SUVmean <- df_baseline$SUVmean

# Add outcome label
forest_toxicity_baseline$DIILD <- "yes"

# ----------------------------------------------------------
# Create dataset for patients without toxicity
# ----------------------------------------------------------

# Load control dataset
df_control <- read_excel("your_control_data.xlsx")
df_control <- data.frame(df_control)

# Split into baseline and interim scans
Time <- df_control$PatientID

df_control_baseline <- df_control %>%
  filter(Iime == "Baseline")

# Select the same features as toxicity dataset
common_columns <- intersect(names(forest_toxicity_baseline),names(df_control_baseline))
forest_no_toxicity_baseline <- df_control_baseline[common_columns]

# Add outcome label
forest_no_toxicity_baseline$DIILD <- "no"

# ----------------------------------------------------------
# Combine datasets
# ----------------------------------------------------------

forest_baseline <- rbind(forest_toxicity_baseline, forest_no_toxicity_baseline)
forest_baseline$DIILD <- as.factor(forest_baseline$DIILD)
str(forest_baseline)

# ----------------------------------------------------------
# Random forest model
# ----------------------------------------------------------

# Define tuning grid
tune_grid <- expand.grid(
  .mtry = 1:(ncol(forest_interim) - 1)
)

# Define number of trees to evaluate
ntree_values <- c(
  50, 100, 150, 200,
  250, 300, 500, 1000
)

# Leave-one-out cross-validation
control <- trainControl(method = "LOOCV")

# rf model
rf_results_baseline <- list()
for (ntree in ntree_values) {

  set.seed(42)

  rf_model_baseline <- train(
    DIILD ~ .,
    data = forest_baseline,
    method = "rf",
    trControl = control,
    tuneGrid = tune_grid,
    ntree = ntree
  )

  rf_model_baseline$ntree <- ntree
  rf_results_baseline[[as.character(ntree)]] <- rf_model_baseline
}

# Select best performing model
best_rf_model_baseline <- rf_results_baseline[[
  which.max(
    sapply(
        rf_results_baseline,
        function(x) max(x$results$Accuracy)
      )
    )
  ]]

cat(
  "Best ntree value:",
  best_rf_model_baseline$ntree,
  "\n"
)

# Plot model performance
plot(best_rf_model_baseline)

# ----------------------------------------------------------
# Feature importance
# ----------------------------------------------------------

final_rf_model_baseline<- best_rf_model_baseline$finalModel
importance_scores_baseline <- importance(final_rf_model_baseline)

importance_df_baseline <- data.frame(
  Feature = rownames(importance_scores_baseline),
  Importance = importance_scores_baseline[
    , "MeanDecreaseGini"
  ]
)

ggplot(importance_df_baseline, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Feature Importance", x = "Feature", y = "Importance (Mean Decrease in Gini)")


# ----------------------------------------------------------
# Multidimensional scaling (MDS) plot
# ----------------------------------------------------------
 
oob_predictions_baseline <- predict(
  final_rf_model_baseline,
  forest_baseline,
  type = "prob"
)[, 2]

distance_matrix_baseline <- dist(oob_predictions_baseline)

mds_result_baseline <- cmdscale(distance_matrix_baseline, k = 2)

mds_df_baseline <- data.frame(
  MDS1 = mds_result_baseline[, 1],
  MDS2 = mds_result_baseline[, 2],
  Actual = forest_baseline$DIILD
)

ggplot(mds_df_baseline, aes(x = MDS1, y = MDS2, color = (Actual))) +
  geom_point() +
  labs(title = "MDS Plot of Random Forest Predictions", 
       x = "MDS Dimension 1", 
       y = "MDS Dimension 2", 
       color = "Group"
       )


```


# 5) Statistics
As the current study is explorative we also wanted to test if any of the remaining features showed significant differences between baseline and interim in patient who developed and who did not develop toxicity during chemotherapy.

As an extra control we also added baseline and interim data from a patient group not receiving the same kind of chemotherapy. Therefore these patients could not develop toxicity at the interim scan.

```{r , echo=FALSE}

# ----------------------------------------------------------
# Testing for significance between baseline and interim
# within patients who developed toxicity
# ----------------------------------------------------------
feature_set_toxicity <- forest_toxicity_baseline %>%
  select(where(is.numeric)) %>%
  names()

forest_toxicity_baseline <- forest_toxicity_baseline %>% arrange(PatientID)
forest_toxicity_interim   <- forest_toxicity_interim %>% arrange(PatientID)

p_values_toxicity <- sapply(feature_set_toxicity, function(feature) {
  
  wilcox.test(
    forest_toxicity_baseline[[feature]],
    forest_toxicity_interim[[feature]],
    paired = TRUE
  )$p.value
})

significance_results_toxicity <- data.frame(
  Feature = feature_set_toxicity,
  P_value = p_values_toxicity
)

significant_feature_names_toxicity <- significance_results_toxicity %>%
  filter(P_value < 0.05) %>%
  pull(Feature)

# ----------------------------------------------------------
# Testing for significance between baseline and interim
# within patients who did not developed toxicity
# ----------------------------------------------------------
feature_set_no_toxicity <- forest_no_toxicity_baseline %>%
  select(where(is.numeric)) %>%
  names()

forest_no_toxicity_baseline <- forest_no_toxicity_baseline %>% arrange(PatientID)
forest_no_toxicity_interim   <- forest_no_toxicity_interim %>% arrange(PatientID)

p_values_no_toxicity <- sapply(feature_set_no_toxicity, function(feature) {
  
  wilcox.test(
    forest_no_toxicity_baseline[[feature]],
    forest_no_toxicity_interim[[feature]],
    paired = TRUE
  )$p.value
})

significance_results_no_toxicity <- data.frame(
  Feature = feature_set_no_toxicity,
  P_value = p_values_no_toxicity
)

significant_feature_names_no_toxicity <- significance_results_no_toxicity %>%
  filter(P_value < 0.05) %>%
  pull(Feature)

# ----------------------------------------------------------
# Testing for significance between baseline and interim
# within control patient group (DLBCL patients)
# ----------------------------------------------------------

df_DLBCL_control <- read_excel("your_control_group_data.xlsx") %>%
  arrange(PatientID)

df_DLBCL_baseline <- df_DLBCL_control %>% filter(Time == "Baseline")
df_DLBCL_interim   <- df_DLBCL_control %>% filter(Time == "Interim")

features_DLBCL <- df_DLBCL_baseline %>%
  select(where(is.numeric)) %>%
  names()

p_values_DLBCL <- sapply(features_DLBCL, function(feature) {
  
  wilcox.test(
    df_DLBCL_baseline[[feature]],
    df_DLBCL_interim[[feature]],
    paired = TRUE
  )$p.value
})

significance_results_DLBCL <- data.frame(
  Feature = features_DLBCL,
  P_value = p_values_DLBCL
)

significant_feature_names_DLBCL <- significance_results_DLBCL %>%
  filter(P_value < 0.05) %>%
  pull(Feature)

```

